Algebraic Groups\vspace{-1ex}

### 2020-01-24 Algebraic Geometry/Algebraic-Geometry.tex
Update

>>>>

\section{January 22, 2020}
Recall that we have the set $\Spec A$ of primes of $A$ and we know the value of the structure sheaf $\O$ on special sets:
\[\O(D(f))=A_f\]
Then we interpolate between these sets to get the value of the sheaf at specific sets.

\subsection{Connection with Locally Ringed Spaces}

\begin{prop}
	Say $(X,\O_X)$ is a LRS. Then the map
	\[\Hom_{LRS}((X,\O_X),\Spec A)\to \Hom_{\Ring}(A,\O_X(X))\]
	where we send $(f,f^\sharp)\mapsto f^\sharp(\Spec A)$, is a bijection.
\end{prop}
\begin{prf}
	Why is this surjective? Given a map $\varphi:A\to \O_X(X)$ we want to concoct a map of schemes $A\to\Spec A$.
	Let $p\in X$ and $f(p)=(\rho_p\circ\varphi)^{-1}(m_p)$ where $\rho_p$ is the localization at $p$ map.

	Why is $f$ continuous under this definition? It is enough to chekc that $f^{-1}(D(a))\subseteq X$ is open,
	which is the set of all $x\in X$ such that $[\varphi(x)]\notin m_x\in \O_{X,x}$ at x. In other words, we are finding the $x$ such that $\overline{\varphi(a)}\ne 0\in k(x)$ (the residue field.)

	Then to get the map $f^\sharp:\O_{\Spec A}\to f_\ast\O_X$.
\end{prf}

\begin{lem}
	Let $(X,\O_X)$ be a locally ringed space. Let $f\in\O_X(X)$. Then 
	\[D(f)=\{x\in X|f\notin m_x\in\O_{X,x}\}\subset X\]
	is open. (Note we are writing the germ $g_x$ as $g$).
\end{lem}
\begin{prf}
	Since $g\notin m_x$, so $g$ is a unit in $\O_{X,x}$. Thus we have an $h\in\O_{X,x}$
	such that $gh=1$. Thus there exists an $x\in U$ and $\tilde h\in \O_X(U)$ such that $g|_U\tilde h=1\in\O_X(U)$
	so $U\subseteq D(g)$.
\end{prf}

\begin{rmk}
	Look at ``sheaf on a base.'' This is the idea of interpolation after defining the sheaf on the distinguished 
	opens $D(f)$. It is in Vakil and EGA chapter zero and many others.
\end{rmk}

\subsection{Schemes}
So what is a scheme?
\begin{defn}
	A \textbf{scheme} is a locally ringed spaces such that there exists an open covering $U_i\subseteq X$ with $(U_i, \O_X(U_i))$ is an affine scheme for all $i$.
\end{defn}
Some examples:
\begin{itemize}
	\item $\Spec \bbZ$. $\bbZ$ is special because it is an initial object in $\Ring$. So for any $X$ we have a 
	canonically-defined map $f:X\to \Spec \bbZ$, where we assign to every point in $X$ a prime number: this is the characteristic of the residue field $k(x)$!
	\item $\Spec k$ where $k$ is a field. Here we have a single point and $\O(\ast)=k$. Then the endomorphisms of this scheme are 
	the endomorphisms $\Hom(k,k)$, which is often interesting! There is a unique one when $k=\bbQ$, but not over $\bbC$!
\end{itemize}

\begin{defn}
	For a fixed scheme $S$, the category of $S$ schemes $\Sch/S$ has objects $X\to S$ and 
	$\Hom(X\to S,Y\to X)$ are given by diagrams.
\end{defn}

\section{January 24th, 2019}
Today we are going to continue talking about $S$-schemes (or schemes over $S$). This is just the slice category $\Sch/S$!
Notice to really make things work we have to make the following definition:
\[(g,g^\sharp)\circ(f,f^\sharp)=(g\circ f,g_\ast(f^\sharp)\circ g^\sharp)\]
\begin{rmk}
	There is a section in Hartshorne where he talks about the difference between $k$-varieties and $k$-schemes. You should read it.
\end{rmk}

\subsection{Examples of Schemes}
Of course we have affine schemes by construction. Other examples we have seen are $\Spec R$ for $R$ a DVR or product of fields. These both 
have two points.

\begin{rmk}
	Look up the ``Punctual Hilbert Scheme'' that talks about schemes with one point with different-sized tangent spaces. For instance, $k[x,y]/(x,y)^2$.
\end{rmk}

Remember the idea of the line with two origins? Well you can take two copies of $\Spec k[[t]]$ (which contains both an open and a closed point) 
and glue along the open point and get something analogous. This is no loner affine, but that is not obvious!

\subsection{Proj}
Let $A=\oplus A_i$ be a graded ring. Then an element of $A$ is homogeneous if it lives in some $A_i$. An ideal is homogeneous if $I$ 
is generated by homogeneous elements.

Some examples: $A=k[x_0,\dots,x_n]$ with $A_i$ the monomials of degree $i$. The irrelevant ideal $A^+$ is the 
ideal of all stuff in positive degree.

Then $\Proj$ makes a schemes from a graded ring. As a set,
\[|\Proj A|=\{\frakp\subseteq A| \frakp\text{ is a homogeneous prim such that } A^+\not\subset\frakp\}\]
and we assign to this set the Zariski topology where if $I\subseteq A$ is homogeneous, then
\[V(I)=\{\frakp\in|\Proj A||\frakp\supset I\}\]
and, in particular, notice that $V(A^+)=\varnothing.$

Then we need to assign to this space a sheaf. Let $f\in A$ be a homogeneous element. Then 
\[\O(D(f))=A_{(f)}=(A_f)_0\]
which is the degree zero part of the usual ring $A_f$.

For example, if $A=k[x_0,x_1]$, and if $f=x_0$, then $A_{(f)}=k\left[\frac{x_1}{x_0}\right]$.

Our main result: 
\begin{thm}
	$(|\Proj A|,\O)$ is a locally ringed space. Furthermore given a homogeneous element $f$ with $\deg f>0$, the map 
	\[A_{(f)}\to\O(D(f))\]
	induces an isomorphism 
	\[(D(f),\O_{D(f)})\xrightarrow{\sim}\Spec A_{(f)}.\]
\end{thm}
Another example: consider $\bbP_\bbZ^n=\Proj\bbZ[x_0,\dots,x_n]$.

### 2020-01-24 Algebraic Groups/Algebraic-Groups.tex
Update

\part{Quarter 2: Representation Theory}
\section{January 17th, 2020}
\subsection{Recap}
Let $T\subseteq B\subseteq G$ be a maximal torus, Borel, and semisimple algebraic group. Then we talked about:
\begin{itemize}
	\item The infinitesimal theory by studying $\Lie G$
	\item Borel subgroups, parabolic subgroups, and the flag variety $G/B$.
	\item We also saw the categorization of split connected semisimple algebraic groups according to their root data:
	This is either given by $(X,R,X^\vee,R^\vee)$ of a group $X$ and a root system $R$ and their duals or equivalently a 
	weight lattice $\Lambda$ and root latice $\Lambda_r=\bbZ R$. Then we can define 
	\[Z/\bbZ R\cong \Lambda/\Lambda_r\]
	to be the \textbf{fundamental group.}
	\item Representation theory: $\Rep G$ is what is called a \textbf{highest weight category.} Let $\Lambda=X$. Then 
	we have the \textbf{dominant roots}
	\[\Lambda_+=\{\lambda\in\Lambda|\langle\lambda,\alpha^\vee\rangle\ge 0,\alpha\in R^+\}\]
	and for every dominant root $\lambda$, we have the \textbf{(co)standard} modules $\Delta(\lambda)$ and $\nabla(\lambda)$.
\end{itemize}
\begin{defn}
	Let $\lambda\in\Lambda=\Hom(T,\Gm)$. Define the $T$-module $\calR_\lambda$ by 
	\[t\cdot 1\eqdef \lambda(t)1.\]
	Then $T(k)=(k^\times)^\Gamma$ gives us $\lambda(t)\in k^\times$.
\end{defn}

@@

>>>>

\part{Quarter 2: Representation Theory}
\section{January 17th, 2020}
\subsection{Recap}
Let $T\subseteq B\subseteq G$ be a maximal torus, Borel, and semisimple algebraic group. Then we talked about:
\begin{itemize}
	\item The infinitesimal theory by studying $\Lie G$
	\item Borel subgroups, parabolic subgroups, and the flag variety $G/B$.
	\item We also saw the categorization of split connected semisimple algebraic groups according to their root data:
	This is either given by $(X,R,X^\vee,R^\vee)$ of a group $X$ and a root system $R$ and their duals or equivalently a 
	weight lattice $\Lambda$ and root latice $\Lambda_r=\bbZ R$. Then we can define 
	\[Z/\bbZ R\cong \Lambda/\Lambda_r\]
	to be the \textbf{fundamental group.}
	\item Representation theory: $\Rep G$ is what is called a \textbf{highest weight category.} Let $\Lambda=X$. Then 
	we have the \textbf{dominant roots}
	\[\Lambda_+=\{\lambda\in\Lambda|\langle\lambda,\alpha^\vee\rangle\ge 0,\alpha\in R^+\}\]
	and for every dominant root $\lambda$, we have the \textbf{(co)standard} modules $\Delta(\lambda)$ and $\nabla(\lambda)$.
\end{itemize}
\begin{defn}
	Let $\lambda\in\Lambda=\Hom(T,\Gm)$. Define the $T$-module $k	_\lambda$ by 
	\[t\cdot 1\eqdef \lambda(t)1.\]
	Then $T(k)=(k^\times)^\Gamma$ gives us $\lambda(t)\in k^\times$.
\end{defn}

@@

\section{January 22nd, 2020}
Now we begin the first part of this class: Infinitesimal theory, derivationts, differentails, and Lie algebras.

\subsection{Tangent Spaces}
If we have a variety $V$, we can define the dual space of $V$ at $x$ by considering 
\[(m_{\O_x}/m^2_{\O_x})^\ast\]
then we can talk about the cotangent space $m/m^2$.

More formally, set $k$ to be a field (note in Jantzen he uses a commutative ring). Let $A$ be a finitely generated \textbf{commutative} $k$-algebra,
so there exists a surjection $k[x_1,\dots,x_n]\to A$. 
\begin{defn}
	Let $M\in\lmod A$. Then a map $D:A\to M$ is a \textbf{derivation} if 
	\[D(ab)=aD(b)=bD(a).\]
	We say that it is a \textbf{$k$-derivation} or $k$-linear if $D(k)=0$. 
\end{defn}

As an example, let $B=k[x_1,\dots,x_n]$ and let $\Omega_B$ be the free $B$ module of rank $n$, generated by $dx_1,\dots,dx_n$.
Then the map $d_B:B\to\Omega_B$ sending $x_i\mapsto dx_i$ and $1\mapsto 0$ is a derivation.
\begin{lem}
	Let $M$ be a $B$ module. Then 
	\[\operatorname{Der}_k(B,M)\stackrel{\sim}{\to}\Hom_B(\Omega_B,M).\]
\end{lem}
\begin{prf}
	Given a derivation $D:B\to M$, we can define a map $\varphi:\Omega_B\to M$ via 
	\[\varphi(dx_i)=D(x_i)\]
	and extending $B$-linearly. One has to prove that this defines an algebra morphism. 
\end{prf}
\begin{rmk}
	Another way to phrase this is $\Omega_B$ is a universal object in $\lmod B$ for derivations.
\end{rmk}
\begin{thm}
	Let $A$ be a finitely generated commutative $k$-algebra. Then there exists a (universal) $A$-module $\Omega_A$
	and a (universal) derivation $d_A:A\to \Omega_A$ such that for any $A$-module $M$,
	\[\operatorname{Der}_k(A,M)\stackrel{\sim}{\to}\Hom_A(\Omega_A,M).\]
\end{thm}
\begin{rmk}
	The universality of the pair $(\Omega_A,d_A)$ is saying that it is unique up to (unique) isomorphism.
\end{rmk}
\begin{prf}
	To see this, we use the last result as well as the fact that 
	\[\Omega_A=\frac{\Omega_B}{I\Omega_B+B(d_B(I))}.\]

	If $I=(f_1,\dots,f_m)$ then $\Omega_A$ is an $A$-module on generators, then $\Omega_A$ is an $A$-module on generators 
	$dx_1,\dots,dx_n$ subject to 
	\[\sum_{i=1}^n\frac{\partial f_j}{\partial x_i}dx_i=0\]
	for all $1\le i\le m$.

	The next part of the proof consists of showing that given a derivation $\tilde D:B\twoheadrightarrow A\xrightarrow{D}M$ and the associated map $\tilde\varphi:\Omega_B\to M$, 
	the latter factors through $\Omega_A$. SHow it vanishes on both part of the sum above.
\end{prf}
\begin{rmk}
	Note: we just proved that the functor $\Der_k:\lmod A\to\Vect_k$ is representable! The uniqueness of the representing object is then a consequence of Yoneda lemma!
\end{rmk}

\begin{ex}
	Let $A=k[x,y]/(x^2+y^2-1)$. Then $\Omega_A$ is generated by $d_x,d_y$ with $2xd_x+2yd_y=0$. When $\ch k=2$,
	we get that $\Omega_A$ is free of rank 2.

	This defines a group scheme (think matrices in $\SL_2\subseteq\GL_2$) and this gives us a surjection of $k[\SL_2]\to A$.

	When $\ch k\ne 2$, we get that $\Omega_A$ is a free $A$-module of rank 1, generated by $dt=xdy-ydx$. I have no idea what this was,
	but she wrote $dy=-xdt.$
\end{ex}
\begin{rmk}
	Notice that the reason we lost smoothness in characteristic 2 was that we lost reducedness: now $(x+y-1)^2=0$. There 
	is a general theorem (see deep in Waterhouse) that an affine group scheme is smooth only if its Hopf algebra is reduced.
\end{rmk}
\subsection{Derivations on Hopf algebras.}
Recall that if $A$ is a Hopf algebra, then $I=\ker\varepsilon$ is the \textbf{augmentation ideal.}
\begin{thm}
	Let $A$ be a finitely generated commutative Hopf algebra over $k$. Then $\Omega_A$ is a free $A$-module.

	In fact, \[\Omega_A\cong A\otimes_k I/I^2\]. The map $d_A:A\to A\otimes I/I^2$ is given by 
	\[a\mapsto a'\otimes \pi(a'')\]
	where the primes indicate Sweedler notation and $\pi:A\to I/I^2$ is a linear map (not a homomorphism!) that takes a (non-unique) 
	splitting $A=I\oplus k\cdot 1$ and then project on the first coordinate and take a quotient by $I^2$. It remains to prove that, once we pass to the 
	quotient, the choice of direct sum complement is irrelevant.
\end{thm}

\begin{lem}
	We say that $D:A\to M$ is an $\varepsilon$-derivation (for $\varepsilon:A\to k$) if 
	\[D(ab)=\varepsilon(a)D(b)+\varepsilon(b)D(a).\]
	The universal module of $\varepsilon$-derivations is 
	\[\Omega_A\otimes_\varepsilon k=\Omega_A/I\Omega_A=I/I^2\]
\end{lem}

\section{January 24th, 2020}
Let $A$ be a finitely generated commutative Hopf algebra over $k$. Often we can drop finite generation 
because any Hopf algebra is a limit of finitely generated ones (because of finite global dimension), so 
these arguments still hold if they can pass to the limit.

Let $\varepsilon:A\to k$ be the counit and $I=\ker\varepsilon$ be the augmentation ideal.
\begin{rmk}
	We have this duality between algebra and geometry, so the counit can be seen as an element of $G_A(k)=\Hom(A,k)$,
	specifically the unit element of the group. :)
\end{rmk}

\begin{lem}
	There is a series of isomorphisms
	\[\Der_\varepsilon(A,M)\simeq \Hom_k(I/I^2,M)\simeq\Hom_A(A\otimes_kI/I^2,M)\]
\end{lem}
This can be proved relatively simply. It is a good exercise.
\begin{thm}
	\begin{itemize}
		\item $\Omega_A\simeq A\otimes_k I/I^2$
		\item $d_A:A\to \Omega_A=A\otimes I/I^2$
	\end{itemize}
	where the map $d_A$ is the one sending $a\mapsto\sum a'\otimes \pi(a'')$
\end{thm}
\begin{prf}
	Recall that $\pi:A\to I/I^2$ is the map that sends $A\cong k\cdot 1\oplus I$ to its projection to $I$ modulo $I^2$.

	Let $M$ be an $A$ module. Give $A\oplus M$ an $A$-algebra structure via 
	\[(a,m)\cdot(a',m')=(aa',am'+a'm).\]
	Then let $G_A(-)=\Hom(A,0)$ and notice 
	\[G_A(A\oplus M)=\Hom_\Alg(A,A\oplus M)\]

	Then we claim that 
	\[\Hom(A,A\oplus M)=\{(\varphi,D)|\varphi\in\Hom(A,A),D\in \Der_\varphi(A,M)\}\]
	then notice that 
	\begin{align*}
		(\varphi,D)(ab) &= (\varphi(a)\varphi(b),\varphi(a)D(b)+D(a)\varphi(b))\\
		&=(\varphi(a),D(a))\cdot(\varphi(b),D(b)).
	\end{align*}
	so this is indeed an algebra morphism. The other direction is clear, so the bijetion holds.

	Now consider the map $p:A\oplus M\to A$ and the induced map
	\[\ker p_\ast\hookrightarrow G_A(A\oplus M)\xrightarrow{G_A(A)}\]
	but the projection map admits a section, so $G_A(A\oplus M)$ is a semidirect product. What is $\ker p_\ast$, though?

	\[\ker p_\ast=\{(\varphi,D)|p_\ast(\varphi,D)=\varphi=\id\in G_A(A)\}\]
	so in particular $\varphi=\varepsilon$. Thus $\ker p_\ast\cong \Der_\varepsilon(A,M)$,
	hence $G_A(A\oplus M)\simeq \Der_\varepsilon(A,M)\rtimes G_A(A)$.

	For all $D'\in \Der_\varepsilon(A,M)$, the element $(D',\id_A)\in\Der_\varepsilon(A,M)\rtimes G_A(A)=G_A(A\oplus M)$
	and this gives us (why?) a ``hidden'' correspondence between the $\varepsilon$ derivations and $k$-linear derivations.

	To find this, take $D'\in\Der_\varepsilon(A,M)$ and consider the element $(\varepsilon, D')\in\ker p_\ast$.
	Multiplying, we get (recalling that multiplication of functions is given by 
	\[(\varphi\cdot\varphi')(x)=m\circ (\varphi\otimes \varphi')\circ \Delta(x)\]
	in a Hopf algebra)
	\[(\varepsilon, D')\cdot(\id_A,0)=(\varepsilon\cdot\id_A,a\mapsto \sum a'D'(a''))\]

	Now use the map $d_A^\varepsilon:A\to A\otimes I/I^2$ via $a\mapsto 1\otimes \pi(a)$ and compute $\id_A\cdot d_A^\varepsilon$, which gives us our map 
	$a\mapsto \sum a'\otimes \pi(a'')$ which proves things apparently but some logic is missing.
\end{prf}

\subsection{Our next goal}
We want to show that, in characteristic zero, any algebraic group scheme is reduced (and thus smooth).
\begin{lem}
	Let $X-1,\dots,x_r$ be a basis for $I/I^2$. Then $\{x_i^{m_i}\cdots x_r^{m_r}|\sum m_i=n\}$
	is a basis for $I^n/I^{n+1}$.
\end{lem}
\begin{rmk}
	Notice that this definitely doesn't work if our algebra is not a Hopf algebra! 
\end{rmk}
\begin{prf}
	Essentially this boils down to the existsnce of derivations with special properties: there is a $D_i:A\to A$ for all $1\le i\le r$
	such that 
	\[D_i(x_j)=\delta_{ij}\pmod{I},\]
	essentially giving us differential operators that distinguish variables.
	To see why the result follows, notice that if this claim is true then the Leibniz rule and induction gets us 
	\[D_r^{m_r}D_{r-1}^{m_{r-1}}\cdots D_1^1(x_1^{m_1}\cdots x_r^{m_r})=m_1!\cdots m_r!\pmod{I}\]
	which implies the statement.

	To generate these operators, we will use the last result. Recall that $\Der_\varepsilon(A,A)=\Hom_k(I/I^2,A)$. Then consider
	\begin{figure}[h]
		\centering
		\begin{tikzcd}
			A\ar[rrd]\ar[d,"\pi"] &  &\\
			I/I^2\ar[r,"d_i"] & k \ar[r,hook] & A
		\end{tikzcd}
	\end{figure}

	\noindent where $d_i(x_j)=\delta_{ij}$. But notice that with such a $d_i$ we get a $k$-derivation $D_i:A\to A$
	defined by 
	\[D_i(a)=\sum a' d_i(\pi(a''))\]
	and 
	\[\varepsilon D_i(a)=\sum \varepsilon(a') d_i(\pi(a''))=d_i(\pi(\sum\varepsilon(a')a''))=d_i(\pi(a))=d_i(\pi(x_j))=\delta_{ij}\]
\end{prf}
